---
title: "STAT 447 Assignment 3"
author: "Caden Hewlett"
date: "`r Sys.Date()`"
header-includes:
   - \usepackage{bbm}
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(extraDistr)
library(ggplot2)
library(pracma)
```


# Question 1 : Functions on the Unit Interval

For this question, use Simple Monte Carlo. The main twist compared to week one is that you will use a continuous random variable.


## Part 1

Write a function called `mc_estimate` that takes a function $f : [0, 1] \rightarrow \mathbb{R}$ and outputs the Monte Carlo estimato of $\int_0^1 f(x)dx$ using $n = 10,000$ independent samples from $\text{unif}(0, 1)$.
 

## Part 2

Consider the function $f: [0, 1] \rightarrow [0, \infty)$ given by:

$$
f(x) = \dfrac{1}{\sqrt[3]{x^2(1- x)}}
$$
Note, importantly, that 

$$
\mathcal{I}_1 =\int_0^1 f(x)\text{d}x = \dfrac{\pi}{\sin\left(\frac{\pi}{3}\right)}
$$
Test your implementation of `mc_estimate` by checking that it produces an answer close to the value above.

```{r equation1result}
eq_1_result = pi / (sin(pi/3))
eq_1_result
```

## Part 3

The following integral, known as the sine integral, does not admit a closed-form expression. 

$$
\mathcal{I}_2 = \int_0^1 \dfrac{\sin(t)}{t}\text{d}t
$$

It does not admit a closed-form expression. Estimate its value using `mc_estimate(f)`.


**Solution**


To test our Monte Carlo Approximation, we will evaluate $\text{Si}(1)$ using the `pracma` package.

```{r six}
Si(1)
```


# Question 2 :  Posterior and Point Estimates

## Part 1

## Part 2 


## Part 3

## Part 4


# Question 3 : Bayes Action



## Part 1


## Part 2
