---
title: "STAT 447 Assignment 9"
subtitle: "MCMC Hacking"
author: "Caden Hewlett"
date: "`r Sys.Date()`"
editor:
  mode: source
header-includes:
   - \usepackage{bbm}
   - \usepackage{upgreek}
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(extraDistr)
library(ggplot2)
library(pracma)
library(distr)
library(latex2exp)
library(knitr)
library(rstan)
library(reticulate)
require(dplyr, quietly = TRUE)
require(rstan, quietly = TRUE)
require(knitr, quietly = TRUE)
```

## Data Import

We import the primary data set inspired by Davidson-Pilon, (2013) below:
```{r import}
# main data vector
sms_data = c(
    13,24,8,24,7,35,14,11,15,11,22,22,11,57,11,19,29,6,19,
    12,22,12,18,72,32,9,7,13,19,23,27,20,6,17,13,10,14,6,
    16,15,7,2,15,15,19,70,49,7,53,22,21,31,19,11,18,20,12,
    35,17,23,17,4,2,31,30,13,27,0,39,37,5,14,13,22)
# data frame
df = data.frame(
  num_texts = sms_data,
  day = 1:length(sms_data)
)
```

## Bayesian Model

We denote $C$ to be the change point, selected uniformly from days $d \in \{1, 2, \dots N\}$  where $N$ is the number of observations. Then, there is a likelihood for days less than change point $C$ and a likelihood for days above the change point.


We can denote the model as follows:
$$
\begin{aligned}
\lambda_{1} &\sim \exp(1/100) \\
\lambda_{2} &\sim \exp(1/100) \\
C &\sim \text{unif}\big(\{1, 2, \dots , N\}\big) \\
Y_d \mid \{C,\lambda_1, \lambda_2\} &\sim \text{pois} \big( \mathbbm{1}[\,d < C\,] \lambda_1 + \mathbbm{1}[\,d \geq C\,]\lambda_2\big)
\end{aligned} 
$$

We provide an implementation of the joint distribution of this model below.

```{r, log_joint}
# inputs are lambdas, C and y
log_joint = function(rates, change_point, y) {
  
  # Return log(0.0) if parameters are outside of the support
  if (rates[[1]] < 0 | rates[[2]] < 0 | change_point < 1 | change_point > length(y)) 
    return(-Inf)
  
  log_prior = 
    dexp(rates[[1]], 1/100, log = TRUE) + 
    dexp(rates[[2]], 1/100, log = TRUE)
  
  log_likelihood = 0.0
  for (i in 1:length(y)) {
    rate = if (i < change_point) rates[[1]] else rates[[2]]
    log_likelihood = log_likelihood + dpois(y[[i]], rate, log = TRUE)
  }
  
  return(log_prior + log_likelihood)
}
```

# Question 1: A Custom MCMC Sampler

## Part 1: Algorithm
Define mathematically an irreducible and invariant MCMC algorithm to sample from the change point model’s posterior.


We will begin by defining two separate kernels $K_{\lambda}$ and $K_{C}$ for the rate parameters $\{\lambda_1,\lambda_2\}$ and the change-point parameter $C$, respectively. We then unify these kernels by defining a kernel mixture for a selection probability $\rho$.


```{r}
# TODO  
```

Now, we consider $K_{C}$. Ideally, we would like to define a symmetric proposal $q_C$; however, we cannot use a Normal Distribution as before so we must find a symmetric discrete distribution. As per the recommendation, this distribution should have variance greater than $3$ to avoid slow mixing.


The simplest possible proposal in this situation would be the discrete $\text{unif}\big( \{a = 1, b =N\} \big)$ distribution, i.e.:
$$
q_C(x' \mid x ) =  \dfrac{1}{b-a+1} = \dfrac{1}{N-1+1} = \dfrac{1}{N}
$$
Where, in this case, $N = 74$. This choice ensures that every possible change point in the support of $C$ can be proposed with nonzero probability.

We can verify by computing the variance of this distribution (to check it will mix nicely):
$$
\text{var}\Big( \text{unif}\, \{a, b\} \mid a = 1, b = N\Big) = \dfrac{(b-a+1)^2 - 1}{12} = \dfrac{(N-1+1)^2 -1}{12} = \dfrac{N^2 - 1}{12} \approx 456 \text{ for } N = 74
$$
Which is certainly a large enough variance to have a breadth of proposal options.



Then, as before, we define the acceptance probability  $\alpha_C(x' \mid x)$ as:
$$
\alpha_C(x' \mid x) = \text{min}\big\{1, r(x', x)\big\} = \text{min}\bigg\{1, \frac{\gamma(x')}{\gamma(x)}\,\bigg\}
$$
We also know that this element of the kernel is irreducible, assuming $\pi(x), \pi(x') > 0$. We know that this is the case because for any arbitrary $x'$, there is a finite number of steps $m$ with nonzero probability required to reach the destination state. In other words, $\mathbb{P}(X^{(m)} = x' | X^{(0)} = x) > 0$. We know that this is the case because the $m$-step transition kernel $K_m =\mathbb{P}(X^{(m)} = x' | X^{(0)} = x) > 0$ is nonzero for all valid $x, x'$ under the restrictions defined earlier.

Let's outline this fact using an arbitrary $K_m(x \mid x')$:
$$
\begin{aligned}
K_m(x \mid x') &= \mathbb{P}(X^{(m+1)} = x' | X^{(m)} = x) \\
K_m(x \mid x') &=  q_C(x' \mid x ) \times \alpha_C(x' \mid x) \\
K_m(x \mid x') &=  \frac{1}{N} \times \text{min}\bigg\{1, \frac{\gamma(x')}{\gamma(x)}\,\bigg\} \\
K_m(x \mid x') &=  \frac{1}{N} \times \text{min}\bigg\{1, \frac{p_{\text{pois}}(x' ; \lambda)}{p_{\text{pois}}(x ; \lambda)}\,\bigg\} \\
\end{aligned}
$$
Where the rate parameter $\lambda$ is simplified from $\mathbbm{1}[\,d < C\,] \lambda_1 + \mathbbm{1}[\,d \geq C\,]\lambda_2 \equiv$ Let's take a moment to verify the term in the MH Ratio. Ideally, $r(x, x')$ should be well-defined and nonzero $\forall (x, x')$ that could be proposed from $q_C(x' \mid x)$. We assume that $x$ is already well-proposed (i.e. $x^{(0)}$ isn't bad). Directly, we know by definition of the Poisson PMF that $\forall x \in \mathbb{Z}^{+}, p_X(x) > 0$. Hence, the ratio is well defined.

Since the ratio is well defined, we know by the above that it will always be the case that $K_m(x \mid x') > 0$, hence all steps are accessible from one another. Thus, $K_C$ does not "break" the irreducibility clause in the MCMC algorithm. 



## Part 2: $\uppi$-Invariance

Prove that the MCMC algorithm you defined in part 1 is $\uppi$-invariant.

Since detailed balance implies global balance, it is sufficient to show that the DBEs hold to show invariance. We know that if each $K_i$ in the kernel mixture is $\uppi$-invariant that their mixture is $\uppi$-invariant. Hence, we will separately prove $\uppi$-invariance for $K_{\lambda}$ and $K_C$ hold under detailed balance to show that $K$ is invariant.





## Part 3: Implementation

Similar to in the slides, we define the kernel function as follows:
```{r}
# true change point at around 25
simulated_yvals = c(round(runif(24, 10, 15)), round(runif(60, 30, 50)))
```
Implement the MCMC algorithm you describe mathematically in `R`. 
```{r q1p3}
N = nrow(df)
# we build such that dim = 1 
mcmc = function(rates, CP, y, n_iterations, debug = FALSE) {
  change_point_trace = rep(-1, n_iterations)
  # initial point
  current_CP = CP
  current_RT = rates
  current = list(rates, CP)
  # iteration station
  for (i in 1:n_iterations) {
    if (debug) {print(unlist(rep("#", times = 10)))}
    if (debug) {print(paste("Iteration:", i))}
    # bernoulli trial
    kernel_choice = ifelse(runif(1) < 0.5, 1, 2)
    if (debug) {print(paste("Chose Kernel", kernel_choice, "..."))}
    # Kernel for Change Point
    if (kernel_choice == 1){
      # Discrete Uniform Proposal
      prop = rdunif(1, 1, N)
      if (debug) {print(paste("Proposed", prop, "..."))}
      # MH ratio
      ratio = (log_joint(current[[1]], prop, y) - 
               log_joint(current[[1]], current[[2]], y))
      # Bernoulli Trial
      if (log(runif(1)) < ratio) {
        # accept
        current[[2]] = prop
        if (debug) {print("Accepted!")}
      } else {
        # reject (redundant but nice)
        current[[2]] = current[[2]]
        if (debug) {print("Rejected!")}
      }
    }
    # Kernel for Lambdas
    else {
      # normal at current point
      ell_1 = rnorm(1, mean = current[[1]][1])
      ell_2 = rnorm(1, mean = current[[1]][2])
      # then the proposal is the vector
      prop = c(ell_1, ell_2)
      if (debug) {print(paste("Proposed", unlist(round(prop,2)), "..."))}
      # MH Ratio
      ratio = (log_joint(prop, current[[2]], y) - 
               log_joint(current[[1]], current[[2]], y))
      if (log(runif(1)) < ratio) {
        # accept
        current[[1]] = prop 
        if (debug) {print("Accepted!")}
      }else {
        # reject (redundant but nice)
        current[[1]] = current[[1]]
        if (debug) {print("Rejected!")}
      }
    # update trace
     change_point_trace[i] = current[[2]]
     if(debug) {Sys.sleep(3)}
    }
  }
  return(
    list(
      change_point_trace = change_point_trace, 
      last_iteration_rates = current[[1]]
    )
  )
}
## ADDITIONAL TESTING 
# test = mcmc(c(0.1, 0.2), 34, simulated_yvals, 200, debug = FALSE)
# plot(test$change_point_trace[test$change_point_trace > 0], type = 'l',
#      ylim = c(1, 74))
# abline(h = 25, col = 'red')
```