---
title: "STAT 447 Assignment 9"
subtitle: "MCMC Hacking"
author: "Caden Hewlett"
date: "`r Sys.Date()`"
editor:
  mode: source
header-includes:
   - \usepackage{bbm}
   - \usepackage{upgreek}
output: pdf_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(extraDistr)
library(ggplot2)
library(pracma)
library(distr)
library(latex2exp)
library(knitr)
library(rstan)
library(reticulate)
require(dplyr, quietly = TRUE)
require(rstan, quietly = TRUE)
require(knitr, quietly = TRUE)
```

## Data Import

We import the primary data set inspired by Davidson-Pilon, (2013) below:
```{r import}
# main data vector
sms_data = c(
    13,24,8,24,7,35,14,11,15,11,22,22,11,57,11,19,29,6,19,
    12,22,12,18,72,32,9,7,13,19,23,27,20,6,17,13,10,14,6,
    16,15,7,2,15,15,19,70,49,7,53,22,21,31,19,11,18,20,12,
    35,17,23,17,4,2,31,30,13,27,0,39,37,5,14,13,22)
# data frame
df = data.frame(
  num_texts = sms_data,
  day = 1:length(sms_data)
)
```

## Bayesian Model

We denote $C$ to be the change point, selected uniformly from days $d \in \{1, 2, \dots N\}$  where $N$ is the number of observations. Then, there is a likelihood for days less than change point $C$ and a likelihood for days above the change point.


We can denote the model as follows:
$$
\begin{aligned}
\lambda_{1} &\sim \exp(1/100) \\
\lambda_{2} &\sim \exp(1/100) \\
C &\sim \text{unif}\big(\{1, 2, \dots , N\}\big) \\
Y_d \mid \{C,\lambda_1, \lambda_2\} &\sim \text{pois} \big( \mathbbm{1}[\,d < C\,] \lambda_1 + \mathbbm{1}[\,d \geq C\,]\lambda_2\big)
\end{aligned} 
$$

We provide an implementation of the joint distribution of this model below.

```{r, log_joint}
# inputs are lambdas, C and y
log_joint = function(rates, change_point, y) {
  
  # Return log(0.0) if parameters are outside of the support
  if (rates[[1]] < 0 | rates[[2]] < 0 | change_point < 1 | change_point > length(y)) 
    return(-Inf)
  
  log_prior = 
    dexp(rates[[1]], 1/100, log = TRUE) + 
    dexp(rates[[2]], 1/100, log = TRUE)
  
  log_likelihood = 0.0
  for (i in 1:length(y)) {
    rate = if (i < change_point) rates[[1]] else rates[[2]]
    log_likelihood = log_likelihood + dpois(y[[i]], rate, log = TRUE)
  }
  
  return(log_prior + log_likelihood)
}
```

# Question 1: A Custom MCMC Sampler

## Part 1: Algorithm
Define mathematically an irreducible and invariant MCMC algorithm to sample from the change point model’s posterior.

## Part 2: $\uppi$-Invariance

Prove that the MCMC algorithm you defined in part 1 is $\uppi$-invariant.

## Part 3: Implementation

Implement the MCMC algorithm you describe mathematically in `R`. 
```{r q1p3}
mcmc = function(rates, change_point, y, n_iterations) {
  change_point_trace = rep(-1, n_iterations)
  
  for (i in 1:n_iterations) {
    # TODO: implement a MCMC sampler
  }
  
  # Return:
  # - the trace of the change points (for question 1) 
  # - the rates at the last iteration (for question 2)
  return(
    list(
      change_point_trace = change_point_trace, 
      last_iteration_rates = rates
    )
  )
}
```